Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-07 Paper/manuscript/manuscript_05.tex
add case studies 4.1 and 4.2. Improve cov. functions in section2

By marginalizing over $\mathbf{f}$ and conditioning on vector of observations $\mathbf{y}$, we obtain the posterior distribution of interest
%
\begin{align}\label{eq:posterior}
	p(f^*|\mathbf{y}) = \frac{\int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)\text{d}\mathbf{f}}{p(\mathbf{y})},
\end{align}
where $p(\mathbf{y})$ is the marginal likelihood and is given by
\begin{align}\label{eq:evidence}
	p(\textbf{y}) = \int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*) \text{d}\bm{f}\text{d}f^*.
\end{align}
%
If the observation model $p(\mathbf{y}|\mathbf{f})$ is Gaussian, both integrals in eq. \eqref{eq:posterior} and \eqref{eq:evidence} can be solved analytically conditioned on the hyperparameters. For example, an isotropic Gaussian likelihood yields the following closed-form solution
\begin{align*}
	p(f^*| \mathbf{y}) = \mathcal{N}(f^*| \textbf{k}_{f^*,\mathbf{f}} (\textbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{f},  k_{f^*,f^*}-\mathbf{k}_{f^*,\mathbf{f}} (\mathbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{k}_{\mathbf{f},f^*}  ),
\end{align*}
where $\sigma^2$ is the noise variance.

While Gaussian processes provide an elegant framework for Bayesian non-parametric regression, there are two problems that often arise in practice. 
in many modelling problems, the observation model is non-Gaussian (e.g. Binomial or Poisson regression) and therefore, the posterior distribution $p(f^*| \mathbf{y})$ is analytically intractable and cannot be obtained in closed-form. Furthermore, it is often of interest to impose a prior distribution on one or more hyperparameters and then marginalize over the uncertainty. This renders the posterior distribution intractable even in the case of a Gaussian likelihood. Approximate inference have to be used to tackle the intractable inference problem, typically based on sampling methods such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011} or distributional approximations such as Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient}. Secondly, due to the matrix inverse in eq.~\eqref{eq:posterior} a direct implementation of Gaussian processes scales cubicly in the number of data points, $\mathcal{O}\left(N^3\right)$,  which can be prohibitively slow for many problems of interest. 
\\
\\
In the following section, we propose an approximate framework for fast and accurate inference for Gaussian processes. Using a basis function expansion, we approximate the Gaussian process with a linear model. This representation has three main advantages: 1) it makes inference considerably faster due to the linear structure, 2) it is simple to implement, which makes it easy to use Gaussian processes as building blocks in more complicated models and can be used as latent function in non-Gaussian observational models allowing modelling flexibility, 3) it can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.
\\
\\
The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\mathbf{x}-\mathbf{x}'$, such that it can be written $k(\mathbf{x},\mathbf{x}') = k(|\mathbf{x}-\mathbf{x}'|)$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(\mathbf{r})$, which means that the covariance is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which, in the unidimensional case (i.e. $\mathbf{x}=x\in \mathbb{R}$, so $\mathbf{r}=r\in \mathbb{R}$), are
%
\begin{align}
k_{\nu=\infty}(r)&=\sigma^2 \exp(-\frac{1}{2} r^2/\ell^2), \nonumber \\
k_{\nu=\frac{1}{2}}(r)&=\sigma^2 \exp(-r/\ell), \nonumber \\
k_{\nu=\frac{3}{2}}(r)&=\sigma^2(1+\sqrt{3}r/\ell) \exp(-\sqrt{3}r/\ell), \nonumber \\
k_{\nu=\frac{5}{2}}(r)&=\sigma^2(1+\sqrt{5}r/\ell+\frac{5}{3}r^2/\ell^2) \exp(-\sqrt{5}r/\ell), \nonumber 
\end{align}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the length-scale and magnitude, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to the multidimensional case, with multidimensional inputs $\mathbf{x}\in \mathbb{R}^D$ and multidimensional length-scale $\boldsymbol{\ell}\in \mathbb{R}^D$ (Gabi: We should derive the multidimensional form of the previous kernels, but I have now doubts how is the multidimensional form for the Matern kernels with $\nu$=1/2, $\nu$=3/2 and $\nu$=5/2). Using a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{align}
S_{\nu=\infty}(w)&= \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \exp\left(-0.5 \ell^2 w^2\right), \\
S_{\nu=\frac{1}{2}}(w)&= 2\sigma^2 \frac{1}{\ell}\left(\frac{1}{\ell^2} + \omega^2\right)^{-1}, \\
S_{\nu=\frac{3}{2}}(w)&= 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}\left(\frac{3}{\ell^2} + \omega^2\right)^{-2}, \\
S_{\nu=\frac{5}{2}}(w)&= \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}\left(\frac{5}{\ell^2} + \omega^2\right)^{-3},  
\end{align}

\noindent where variable $w$ is in the frequency domain. The $D$-dimensional spectral density functions associated with the Mat\'ern covariance functions considered in the previous section are
%
\begin{align}
S^{\ast}_{\nu=\infty}(\mathbf{w})&= \sigma^2 \sqrt{2\pi}^D \prod_{d=1}^D l_d \cdot \exp\left(-0.5 \sum_{d=1}^D l_d^2 \mathbf{w}_d^2\right), \nonumber \\
%
S^{\ast}_{\nu}(\mathbf{w})&= \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\mathbf{w}^2 \right)^{(\nu+D/2)}, \nonumber
\end{align}

\noindent where the $D$-vector $\mathbf{w}$ is in the frequency domain.

@@

The finite number $m$ of basis functions in the approximation needed to satisfy eq. (\ref{diff_covs}) depends on the non-linear effects of the function to be learned, that is, on its lengthscale $\ell$. The approximation also depends on boundary $L$ (equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})) which will affect the performance specially near the boundaries. The boundary will also have an effect on the number of basis functions needed in the approximation. The boundary $L$ is set as an extension of the desired predicting input domain $\Psi$. Considering a zero-mean input domain $\Psi=[-S,S] \in \mathbb{R}$, the boundary is defined as
%
\begin{equation}
L=c \cdot S,
\end{equation}

@@

This example consist of a simulated dataset with 250 draws values from a Gaussian process prior with a Matern covariance function (a Matern with 3/2 degrees of freedom) and inputs $\{\mathbf{x}_i\}_{i=1}^{250} \in \{-1,1\} \subset {\rm I\!R}$. The hyperparameters, marginal variance $\alpha$ and lengthscale $\ell$, of this GP prior were set to 1 and 0.15, respectively. The Gaussian noise $\sigma$ added to these draws was set to 0.2.

The regular GP model for fitting this simulated dataset $\mathbf{y}$ can be written as follows,
%
\begin{eqnarray}\label{eq:latentgp_simudata1}
\begin{split}
\mathbf{y} &= f(\mathbf{x}) + \boldsymbol{\epsilon} \\
\boldsymbol{\epsilon} &\sim \mathcal{N}(0, \sigma^2  \mathbf{I}) \\
f(\mathbf{x}) &\sim \mathcal{GP}(0, k(\mathbf{x}, \mathbf{x}', \theta)),
\end{split}
\end{eqnarray}

\noindent The previous formulation corresponds to the latent form where $f(\mathbf{x})$ represents the functional model underlying the noisy data and modeled by a GP prior with a Matern covariance function $k$ (4). $\mathbf{I}$ represents the identity matrix. Saying that the function values $f(\mathbf{x})$ follows a GP model is equivalent to say that $f(\mathbf{x})$ are multivariate Gaussian distributed with covariance matrix $K$, where the element $K_{ij}$ is formed by evaluating the covariance function $k(\mathbf{x}_i,\mathbf{x}_j,\theta)$ at input values $\mathbf{x}_i$ and $\mathbf{x}_j$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using sampling methods such as HMC, would be the marginalized form of the latent GP model in (\ref{eq:latentgp_simudata1}),
%
\begin{equation*}\label{eq:marginalizedgp_simudata1}
\mathbf{y} \sim \mathcal{N}(0, K + \sigma^2 \mathbf{I} )
\end{equation*}

@@

In the HSGP model, the covariance function $k$ is approximated as (\ref{approxcov}),
%
\begin{equation*}
k(x,x') \approx \sum_{j}^{m}S(\sqrt{\lambda_j}) \phi_j(x) \phi_j(x'),
\end{equation*}

\noindent with the spectral density $S$ (3),
%
\begin{eqnarray*}
S(w)&=& 4\alpha^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \omega^2)^{-2},  
\end{eqnarray*}

\noindent and the latent function values $f(\mathbf{x)}$ are approximated as (\ref{approxf}),
%
\begin{equation*}
f(x) \approx \sum_{j}^m \left( S(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j, \nonumber
\end{equation*}

@@

For the HSGP model we use $m=80$ basis functions and a boundary factor $c=1.2$. For the splines model we use 80 knots or basis in the model.

Figure \ref{fig10_Posteriors_exI} shows the posteriors distributions of the three models, the regular GP, the HSGP and the splines, jointly with the true data-generative function and the noisy observations. The right and left extremes and three small regions around the middle of the plot correspond to out-of-sample or test data which have not been taking part on training the models. The test data have been plotted in Figure \ref{fig10_Posteriors_exI} with crosses and the training data with circles. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation.

Figure \ref{fig11_MSE_exI_inter} shows the  standardized root mean squared error (SRMSE), for interpolation and extrapolating data, as a function of the number of basis functions and knots. For this, different models with different number of basis functions for the HSGP model and different number of knots for the splines model have been formulated and fitted on this data. The SRMSE is computed against the true function. Model interpolation has been assessed either using the test data for interpolating and the training data. From Figures \ref{fig10_Posteriors_exI} and \ref{fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the regular GP model, for interpolating and extrapolating data. However, the splines model does not extrapolate data properly. For interpolation both models show roughly similar performance.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.350]{fig11_MSE_exI_inter.pdf}}
\subfigure{\includegraphics[scale=0.350]{fig11_MSE_exI_extra.pdf}}
\caption{Standardized mean square error (SMSE) of the different methods against the true function. (left) SMSE for interpolation. (right) SMSE for extrapolation}
  \label{fig11_MSE_exI_inter}
\end{figure}

\subsection{Gay data}\label{sec:gay_data}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 binomial observations per age group $i$ ($i=1,\cdots,74$), with the observation of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group.

The observational model for the observed number of people $y_i$ supporting same-age marriage per age group $i$ is a binomial model with the total number of people $n_i$ per age group and the probability $p_i$ per age group as model parameters,
%
\begin{equation*}
y_i \sim \text{Binomial}(p_i, n_i).
\end{equation*}

\noindent The total number of people per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ per age group $i$ or mean number of support people per age group $i$. The probability $p_i$ is related to the Gaussian process latent values $f(\mathbf{x})$ with age values $\mathbf{x}=\{x_i\}_{i=1}^{74}$ as inputs, through the logit function,
%
\begin{eqnarray*} \label{eq:gpprior_gay}
p_i &=& \text{logit}(f(x_i)) \nonumber \\
f(\mathbf{x}) &\sim& \mathcal{GP}(0, k(\mathbf{x},\mathbf{x}', \theta)
\end{eqnarray*}

\noindent with a squared exponential covariance function $k$ (2).

In the HSGP model the covariance function is approximated as in eq. (\ref{approxcov}) with the square exponential spectral density in eq. (6), and the latent function $f(\mathbf{x})$ as in eq. (\ref{approxf}).

For the HSGP model we use $m=20$ basis functions and a boundary factor $c=1.5$. For the splines model we use 20 knots or basis in the model.

Figure \ref{fig12_Posteriors_gaydata} shows the posteriors distributions of the three models, the regular GP, the HSGP and the splines, jointly with the true data-generative function and the noisy observations. The HSGP model posterior shown in the figure corresponds to 20 basis functions and a boundary factor of 1.5. For the spline posterior shown in the figure, it correspond to a 20 knots. Two small sets of observations have been used as test data, plotted as crosses in the Figure.

@@

For the HSGP model, different models with different number of basis functions and boundary factors have been used. The root mean square errors (RMSE) for every one of these models have been computed and plotted as a function of the number of basis functions and the boundary factor in Figure \ref{fig13_MSE_train_BF_gaydata}, for training (left) and test (right) data. In this case, the RMSE has been computed against the regular GP model. The expected patterns of the approximation in function of the number of basis function and boundary factor are recognized: as the boundary factor increases, more basis functions are needed.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.40]{fig13_MSE_train_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40]{fig13_MSE_pred_BF_gaydata.pdf}}
\caption{Mean square error (MSE) of the different methods against the true function. (left) MSE for training data. (right) MSE for test data.}
  \label{fig13_MSE_train_BF_gaydata}
\end{figure}

Figure \ref{fig14_MSE_train_gaydata} shows the RMSE of the different models, regular GP, HSGP and splines, against the actual data, for training and test data. Again we can see how the splines models does not extrapolate data properly.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.40]{fig14_MSE_train_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40]{fig14_MSE_pred_gaydata.pdf}}
\caption{Mean square error (MSE) of the different methods against the true function. (left) MSE for training data. (right) MSE for test data.}
  \label{fig14_MSE_train_gaydata}
\end{figure}

@@

\vspace{2mm}
\section*{Acknowledgment}

>>>>

By marginalizing over $\mathbf{f}$ and conditioning on vector of observations $\mathbf{y}$, we obtain the posterior distribution of interest
%
\begin{align}\label{eq:posterior}
	p(f^*|\mathbf{y}) = \frac{\int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)\text{d}\mathbf{f}}{p(\mathbf{y})},
\end{align}
where $p(\mathbf{y})$ is the marginal likelihood and is given by
%
\begin{align}\label{eq:evidence}
	p(\textbf{y}) = \int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*) \text{d}\bm{f}\text{d}f^*.
\end{align}
%
If the observation model $p(\mathbf{y}|\mathbf{f})$ is Gaussian, both integrals in eq. \eqref{eq:posterior} and \eqref{eq:evidence} can be solved analytically conditioned on the hyperparameters. For example, an isotropic Gaussian likelihood yields the following closed-form solution
\begin{align*}
	p(f^*| \mathbf{y}) = \mathcal{N}(f^*| \textbf{k}_{f^*,\mathbf{f}} (\textbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{f},  k_{f^*,f^*}-\mathbf{k}_{f^*,\mathbf{f}} (\mathbf{K}_{\mathbf{f},\mathbf{f}} + \sigma^2 \mathbf{I})^{-1} \mathbf{k}_{\mathbf{f},f^*}  ),
\end{align*}
where $\sigma^2$ is the noise variance.

While Gaussian processes provide an elegant framework for Bayesian non-parametric regression, there are two problems that often arise in practice. 
in many modelling problems, the observation model is non-Gaussian (e.g. Binomial or Poisson regression) and therefore, the posterior distribution $p(f^*| \mathbf{y})$ is analytically intractable and cannot be obtained in closed-form. Furthermore, it is often of interest to impose a prior distribution on one or more hyperparameters and then marginalize over the uncertainty. This renders the posterior distribution intractable even in the case of a Gaussian likelihood. Approximate inference have to be used to tackle the intractable inference problem, typically based on sampling methods such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011} or distributional approximations such as Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient}. Secondly, due to the matrix inverse in eq.~\eqref{eq:posterior} a direct implementation of Gaussian processes scales cubicly in the number of data points, $\mathcal{O}\left(N^3\right)$,  which can be prohibitively slow for many problems of interest. 
\\
\\
In the following section, we propose an approximate framework for fast and accurate inference for Gaussian processes. Using a basis function expansion, we approximate the Gaussian process with a linear model. This representation has three main advantages: 1) it makes inference considerably faster due to the linear structure, 2) it is simple to implement, which makes it easy to use Gaussian processes as building blocks in more complicated models and can be used as latent function in non-Gaussian observational models allowing modelling flexibility, 3) it can be made arbitrary accurate and the trade-off between computational complexity and approximation accuracy can easily be controlled.
\\
\\
The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\bm{\tau}=\mathbf{x}-\mathbf{x}' \in \mathbb{R}^D$, such that it can be written $k(\mathbf{x},\mathbf{x}') = k(\bm{\tau})$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(r), r\in \mathbb{R}$, which means that the covariance is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which, in the unidimensional case (i.e. $\mathbf{x}=x\in \mathbb{R}$, so $\mathbf{r}=r\in \mathbb{R}$, and $\ell\in \mathbb{R}$ ), are
%
\begin{align*}
k_{\nu=\infty}(r)&=\sigma^2 \exp(-\frac{1}{2} r^2/\ell^2), \\
k_{\nu=\frac{1}{2}}(r)&=\sigma^2 \exp(-r/\ell),  \\
k_{\nu=\frac{3}{2}}(r)&=\sigma^2(1+\sqrt{3}r/\ell) \exp(-\sqrt{3}r/\ell), 
%k_{\nu=\frac{5}{2}}(r)&=\sigma^2(1+\sqrt{5}r/\ell+\frac{5}{3}r^2/\ell^2) \exp(-\sqrt{5}r/\ell), 
\end{align*}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the length-scale and magnitude, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to a $D$-dimensional case, by using the property of separability of multidimensional kernels. Let $\mathbf{x}\in \mathbb{R}^D$, $\mathbf{\bm{\tau}}=\{\tau_1,\tau_2\}\in \mathbb{R}^D$ and $\boldsymbol{\ell}=\{\ell_1,\ell_2\}\in \mathbb{R}^D$,
%
\begin{align*}
k_{\nu=\infty}(\bm{\tau})&=\exp(-\frac{1}{2} \tau_1^2/\ell_1^2) \cdot \exp(-\frac{1}{2} \tau_2^2/\ell_2^2), \\
k_{\nu=\frac{1}{2}}(\bm{\tau})&=\exp(-\tau_1/\ell_1) \cdot \exp(-\tau_2/\ell_2), \\
k_{\nu=\frac{3}{2}}(\bm{\tau})&=(1+\sqrt{3}\tau_1/\ell_1) \exp(-\sqrt{3}\tau_1/\ell_1) \cdot (1+\sqrt{3}\tau_2/\ell_2) \exp(-\sqrt{3}\tau_2/\ell_2) \\
\end{align*}

\noindent Using a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. In the unidimensional case the spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{align}\label{eq:specdens}
S_{\nu=\infty}(w)&= \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \exp\left(-0.5 \ell^2 w^2\right), \\
S_{\nu=\frac{1}{2}}(w)&= 2\sigma^2 \frac{1}{\ell}\left(\frac{1}{\ell^2} + \omega^2\right)^{-1}, \\
S_{\nu=\frac{3}{2}}(w)&= 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}\left(\frac{3}{\ell^2} + \omega^2\right)^{-2}, \\
%S_{\nu=\frac{5}{2}}(w)&= \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}\left(\frac{5}{\ell^2} + \omega^2\right)^{-3},  
\end{align}

\noindent where variable $w\in \mathbb{R}$ is in the frequency domain, and $\ell\in \mathbb{R}$. By using the separability property of the spectral densities functions, the $D$-dimensional form of these spectral density functions, where $\mathbf{w}=\{w_1,w_2\}$ and $\mathbf{\ell}=\{\ell_1,\ell_2\}$ are
%
\begin{align}\label{eq:specdens_multi}
S_{\nu=\infty}(\bm{\omega})&= \sqrt{2\pi} \cdot \ell_1 \cdot \exp\left(-0.5 \ell_1^2 \omega_1^2\right) \cdot  \sqrt{2\pi} \cdot \ell_2 \cdot \exp\left(-0.5 \ell_2^2 \omega_2^2\right) \\
S_{\nu=\frac{1}{2}}(\bm{\omega})&= 2 \frac{1}{\ell_1}\left(\frac{1}{\ell_1^2} + \omega_1^2\right)^{-1} \cdot 2 \frac{1}{\ell_2}\left(\frac{1}{\ell_2^2} + \omega_2^2\right)^{-1} \\
S_{\nu=\frac{3}{2}}(\bm{\omega})&= 4 \frac{\sqrt{3}}{\ell_1}^{3}\left(\frac{3}{\ell_1^2} + \omega_1^2\right)^{-2} \cdot 4 \frac{\sqrt{3}}{\ell_2}^{3}\left(\frac{3}{\ell_2^2} + \omega_2^2\right)^{-2},
\end{align}

\noindent which can also be derived from the following expressions \citep{rasmussen2006gaussian}
%
\begin{align*}
S^{\ast}_{\nu=\infty}(\mathbf{\omega})&= \sigma^2 \sqrt{2\pi}^D \prod_{d=1}^D l_d \cdot \exp\left(-0.5 \sum_{d=1}^D l_d^2 \mathbf{\omega}_d^2\right), \\
%
S^{\ast}_{\nu}(\mathbf{\omega})&= \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\mathbf{\omega}^2 \right)^{(\nu+D/2)}.
\end{align*}

@@

The finite number $m$ of basis functions in the approximation needed to satisfy eq. (\ref{diff_covs}) depends on the non-linear effects of the function to be learned, that is, on its lengthscale $\ell$. The approximation also depends on boundary $L$ (equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})) which will affect the performance specially near the boundaries. The boundary will also have an effect on the number of basis functions needed in the approximation. The boundary $L$ is set as an extension of the desired predicting input domain $\Psi$. Considering a zero-mean input domain $\Psi=[-S,S] \in \mathbb{R}$, the boundary is defined as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

This example consists of a simulated dataset with 250 draws of a Gaussian process prior with a Mat{\'e}rn covariance function with 3/2 degrees of freedom and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, and inputs values $\{\mathbf{x}_i\}_{i=1}^{250} \in \{-1,1\} \subset {\rm I\!R}$. To form the final dataset $\mathbf{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.

The regular GP model for fitting this simulated dataset $\mathbf{y}$ can be written as follows,
%
\begin{eqnarray*}\label{eq:latentgp_simudata1}
\begin{split}
\mathbf{y} &= f(\mathbf{x}) + \boldsymbol{\epsilon} \\
\boldsymbol{\epsilon} &\sim \mathcal{N}(0, \sigma^2  \mathbf{I}) \\
f(\mathbf{x}) &\sim \mathcal{GP}(0, k(\mathbf{x}, \mathbf{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $\mathbf{I}$ represents the identity matrix. The previous formulation corresponds to the latent form of a GP model, where $f(\mathbf{x})$ represents the underlying functional model to the noisy data, and it is modeled by a GP prior with a Mat{\'e}rn($\nu=3/2$) covariance function $k$ (\ref{covfun}). Saying that the function values $f(\mathbf{x})$ follows a GP model is equivalent to say that $f(\mathbf{x})$ are multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(\mathbf{x}_i,\mathbf{x}_j,\theta)$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using sampling methods such as HMC, would be its marginalized form,
%
\begin{equation*}\label{eq:marginalizedgp_simudata1}
\mathbf{y} \sim \mathcal{N}(0, K + \sigma^2 \mathbf{I} )
\end{equation*}

@@

In the HSGP model, the latent function values $f(x)$ are approximated as (\ref{approxf}),
%
\begin{equation*}
f(x) \approx \sum_{j}^m \left( S(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j,
\end{equation*}

\noindent with the spectral density $S$ (\ref{specdens}) as a function of $\sqrt{\lambda_j}$,
%
\begin{eqnarray*}
S(\sqrt{\lambda_j})&=& 4\alpha^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \lambda_j)^{-2}, 
\end{eqnarray*}

\noindent and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ (\ref{bf&lambda})
%
\begin{eqnarray*}
\begin{split}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right).
\end{split}
\end{eqnarray*}

\noindent $L$ is the boundary and $m$ the number of basis functions. The parameters $\beta_j$ are $\mathcal{N}(0,1)$ distributed, and $\alpha$ and $\ell$ are the marginal variance and lengthscale parameters of the approximate covariance function, respectively.

In order to do model comparison, in addition to the regular GP model and HSGP model, an splines-based model is also fitted using the Thin Plate Regression Splines approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms}.

Figure \ref{fig10_Posteriors_exI} shows the posteriors distributions of the three models, the regular GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$ (\ref{eq:boundary})), and the splines model with 80 knots. The true data-generative function and the noisy observations are also plotted in the Figure. The right and left extremes and three small regions in the middle of the plot correspond to out-of-sample or test data which have not been taking part on training the models. The test data have been plotted as crosses and the training data as circles. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation.

In order to assess performance of the models as a function the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model and different number of knots for the splines model have been fitted. Figure \ref{fig11_MSE_exI_inter} shows the  standardized root mean squared error (SRMSE), for interpolation and extrapolating data, as a function of the number of basis functions and knots. The SRMSE is computed against the true function. Model interpolation has been assessed either using the test data for interpolating and the training data. From Figures \ref{fig10_Posteriors_exI} and \ref{fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the regular GP model, for interpolating and extrapolating data. However, the splines model does not extrapolate data properly. For interpolation both models show roughly similar performance.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.350]{fig11_MSE_exI_inter.pdf}}
\subfigure{\includegraphics[scale=0.350]{fig11_MSE_exI_extra.pdf}}
\caption{Standarized mean square error (SMSE) of the different methods againts the true function. (left) SMSE for interpolation. (right) SMSE for extrapolation}
  \label{fig11_MSE_exI_inter}
\end{figure}

\subsection{Gay data}\label{sec:gay_data}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 observations of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group $i$ ($i=1,\cdots,74$). The observational model is a binomial model with parameters population $n_i$ and probability of supporting same-sex marriage $p_i$ per age group $i$,
%
\begin{equation*}
y_i \sim \text{Binomial}(p_i, n_i).
\end{equation*}

\noindent The population per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ or mean number of support people per age group. The probabilities $\mathbf{p}=\{p_i\}_{i=1}^{74}$ are modeled by a GP latent function $f(\mathbf{x})$ with a squared exponential covariance function $k$ (2) and age input values $\mathbf{x}=\{x_i\}_{i=1}^{74}$, with the {\it logit} function as a link function,
%
\begin{eqnarray*} \label{eq:gpprior_gay}
p_i &=& \text{logit}(f(x_i)) \nonumber \\
f(\mathbf{x}) &\sim& \mathcal{GP}(0, k(\mathbf{x},\mathbf{x}', \theta).
\end{eqnarray*}

In the HSGP model the latent function $f(\mathbf{x})$ is approximated as in eq. (\ref{approxf}), with the square exponential spectral density as in eq. (\ref{specdens}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in eq. (\ref{bf&lambda}).

@@

Figure \ref{fig12_Posteriors_gaydata} shows the posteriors distributions of the three models, the regular GP, the HSGP model with $m=20$ basis functions and boundary factor $c=1.5$, and the splines model with 20 knots. The true data-generative function and the noisy observations are also plotted in the Figure. Two small sets of observations have been used as test data, and plotted as crosses in the Figure.

@@

For the HSGP model, different models with different number of basis functions and boundary factors have been fitted. The root mean square errors (RMSE) for every one of these models have been computed and plotted as a function of the number of basis functions and the boundary factor in Figure \ref{fig13_MSE_train_BF_gaydata}, for training (left) and test (right) data. In this case, the RMSE has been computed against the regular GP model. The expected patterns of the approximation as a function of the number of basis functions and boundary factor are recognized: as the boundary factor increases, more basis functions are needed.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.40]{fig13_MSE_train_BF_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40]{fig13_MSE_pred_BF_gaydata.pdf}}
\caption{Mean square error (MSE) of the different methods againts the true function. (left) MSE for training data. (right) MSE for test data.}
  \label{fig13_MSE_train_BF_gaydata}
\end{figure}

Figure \ref{fig14_MSE_train_gaydata} shows the RMSE of the different models, regular GP, HSGP and splines, against the actual data, for training and test data. We can see how the splines models does not extrapolate data properly.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.40]{fig14_MSE_train_gaydata.pdf}}
\subfigure{\includegraphics[scale=0.40]{fig14_MSE_pred_gaydata.pdf}}
\caption{Mean square error (MSE) of the different methods againts the true function. (left) MSE for training data. (right) MSE for test data.}
  \label{fig14_MSE_train_gaydata}
\end{figure}

@@

\section{Multidimensional generalization of covariance functions and spectral densities}

\begin{itemize}
\item 
$x, x' \in \mathbb{R}$,\\
$\ell \in \mathbb{R}$, \\
$\tau=x-x'\in \mathbb{R}$, \\
$r=||x-x'||\in \mathbb{R}$, \\
$\tau=r$ \\
$\omega\in \mathbb{R}$
